MCML with Boundary Surfaces

### 2019-06-13 docs/report/article.tex
Update report

\documentclass[]{article}
\usepackage[english]{babel}
\usepackage{url}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{listings}
\bibliographystyle{IEEEtranS}

@@

@@

\section{Implementation}

@@

\subsection{Boundary Specification}

@@

\subsection{Boundary Intersection Test}

@@

\begin{itemize}
\item The binary output is currently not supported. MCO files are always written as ASCII.
\item Glass layers are not supported. Reflectance will always be reported as diffuse.
\item The maximum number of photons is limited to $2^{32}-1$. Also the number of photons must be higher than the number of GPU threads, which will be printed on program startup.
\item Input data is not checked for sanity.
\item All calculations are preformed with 32-bit floating point precision. This is also the case for CUDAMCML and was shown to have only slight effect.
\end{itemize}

@@

\autoref{verification1} compares the diffuse reflectance profiles of clustermcml and CUDAMCML using 100 million photons and one layer of thickness $dl=0.1cm$. The absorption coefficient is $\mu_a=0.1$, the scattering coefficient $\mu_s=1$ and scattering anisotropy $g=0.5$. The refractive indices are $n=1.34$ inside the layer and $1$ outside. The detection window has a size of $1cm$ and a resolution of $0.005cm$. Note that a logarithmic scale is used for a quantity, that is reflectance multiplied by radius. The plots indicate correct results despite different amounts of noise.

@@

@@

@@

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/result2.pdf}
	\caption{Comparison of diffuse reflectance profiles of flat surface with 4 variations of curved surfaces.}
	\label{result2}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{img/result3.pdf}
\caption{Comparison of diffuse reflectance profiles of flat surface with 2 variation of noise surfaces.}
\label{result3}
\end{figure}

@@

\bibliography{bibliography}
\end{document}

>>>>

\documentclass[]{article}
\usepackage[english]{babel} % advanced multi-language features
\usepackage{graphicx} % figures
\usepackage{hyperref} % references and URLs are clickable
\usepackage{listings} % simple code listings (monospaced font, line oriented, no syntax highlighting)
\usepackage[section]{placeins} % restricting floating elements like figures with FloatBarriers
\bibliographystyle{IEEEtranS}

@@

Specifics with regard to multiple layers, variable boundary shapes and general implementation are described in \autoref{impl}.

@@

\section{Implementation}
\label{impl}

@@

\subsection{Boundary Specification}
\label{impl:boundary_specification}

@@

\subsection{Boundary Intersection Test}
\label{impl:intersection}

@@

\begin{itemize}
\item The binary output is currently not supported. MCO files are always written as ASCII.

\item Glass layers are not supported. Reflectance will always be reported as diffuse.

\item The maximum number of photons is limited to $2^{32}-1$. Also the number of photons must be higher than the number of GPU threads, which will be printed on program startup.

\item Interaction properties are not checked for sanity or physical plausibility.

\item All calculations are preformed with 32-bit floating point precision. This is also the case for CUDAMCML and was shown to have only slight effects, that can be neglected.
\end{itemize}

@@

@@

The purpose of this section is to verify, that the results of the new MCML implementation are correct. To this end, CUDAMCML serves as reference, which was already verified against the original MCML. Generally, because of the stochastic nature of the Monte Carlo method, one cannot compare exact outputs. Even if the same seed and same PRNG were used, results are not deterministic because the order of execution of the threads is undefined. Therefore, radiance profiles of both implementations are compared by plotting them into the same coordinate system.

The focus of this project lies on the 2D data of reflectance and transmittance as function of radius. Therefore, and because these profiles are easier to compare than 3D data, verification is confined to them.

It is expected, that the plots differ due to noise, but important characteristics of the curves should be the same. One important characteristic arises from shallow layers. Considering single scattered photons, escape at a certain radius on the top boundary becomes impossible, because the layer depth restricts the escape angles at that point to be grazing angles, i.e. angles of total reflection. Below that radius, a majority of escaping photons are single scattered, with the number of photons falling fast, the less straight they pass (governed by Fresnel formula). Above that radius, escaping photons are multiple scattered, with mixed escape angles, causing the falling rate to be governed by the spatial distribution of photons weight. The change in slope is clearly visible in plots, of which one example is presented next.

\autoref{verification1} compares the diffuse reflectance profiles of clustermcml and CUDAMCML using 100 million photons and one layer of thickness $dl=0.1cm$. The absorption coefficient is $\mu_a=0.1$, the scattering coefficient $\mu_s=1$ and scattering anisotropy $g=0.5$. The refractive indices are $n=1.34$ inside the layer and $1$ outside. The detection window has a size of $1cm$ and a resolution of $0.005cm$, i.e. using $200$ bins. \autoref{verification2} compares transmission profiles of the same simulation.

Plotted on the y-axis is a quantity, that is radiance multiplied by radius. The quantity is chosen that way to account for the larger area at large radii. Note that radiance as recorded by MCML is probability per unit area. Also note that a logarithmic scale is used.

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/verification2.pdf}
	\caption{Corresponding transmission profiles for~\autoref{verification1}.}
	\label{verification2}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{img/verification3.pdf}
\caption{Diffuse reflectance profiles for a thicker layer than the one in~\autoref{verification1}.}
\label{verification3}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{img/verification4.pdf}
\caption{Corresponding transmission profiles for~\autoref{verification3}.}
\label{verification4}
\end{figure}

\autoref{verification3} and~\autoref{verification4} compare profiles for a thicker layer of $dl=1cm$. The main curve characteristic here is smoothness. Through the low amount of noise, the curves of both implementations almost exactly match each other.

The plots indicate correct results despite different amounts of noise. It should be noted that CUDAMCML uses a different PRNG (Multiply-with-Carry), which can affect noise.

Verification was only possible with flat boundary surfaces. However, to provide validity to some degree with respect to structured boundaries as described in~\autoref{impl:boundary_specification} and~\autoref{impl:intersection}, the same intersection algorithm was used, while input data consisted of degenerated "flat" cones.

@@

The parameters can be read from the legend, which uses the same naming convention as in the previous section. Boundary surfaces are shown as dotted lines in the same color as the corresponding radiance profile. The following shapes were used.

\begin{description}
	\item[flat50] A trivial heightfield with all heights set to zero. As all other shapes, it is defined by $50$ samples with uniform spacing of $0.01cm$.
	\item[hihill] An upward curve with high curvature.
	\item[lohill] An upward curve with low curvature.
	\item[hicrater] A downward curve with high curvature.
	\item[locrater] A downward curve with low curvature.
	\item[hinoise] Noise with high amplitude.
	\item[lonoise] Noise with low amplitude.
\end{description}

The shapes are the same at the top and bottom boundary of the layer.

The heightfields are centered around an area-weighted mean height, to be comparable with the flat boundary. The weighting is done, because, in the radial model, outer height values contribute to a larger part of the surface. The mean is computed by

\begin{equation}
	h_m = \frac{\sum_{i=0}^{N}h_i * 2\pi r_i}{\sum_{i=0}^{N}2\pi r_i}
\end{equation}

where $N$ is the number of samples, $h_i$ is the height at index $i$ and $r_i$ is the radius, i.e. the sum of spacings up to index $i$.

@@

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/result2.pdf}
	\caption{Comparison of diffuse reflectance profiles of flat surface with 4 variations of curved surfaces. Solid lines belong to the left and dotted lines to the right axis.}
	\label{result2}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{img/result3.pdf}
\caption{Comparison of diffuse reflectance profiles of flat surface with 2 variation of noise surfaces. Solid lines belong to the left and dotted lines to the right axis.}
\label{result3}
\end{figure}

@@

\FloatBarrier
\bibliography{bibliography}
\end{document}
